---
title: "Estimating and filtering out ambient mRNA for sc-/sn-RNA-Seq Analysis in 10X Flex Genomics data"
author: "Antonia Chroni for SJCRH DNB_BINF_Core"
papersize: a4
fontsize: 11pt
links-as-notes: true
output:
  html_document:
    toc: TRUE
    toc_float: TRUE
    df_print: paged
    code_folding: hide
    toc_depth: 2
    highlight: tango
    number_sections: TRUE
  pdf_document:
    toc: TRUE
    highlight: tango
    number_sections: TRUE
    latex_engine: lualatex
    keep_tex: FALSE
    fig_caption: yes
    fig_crop: no
    fig_height: 2
    fig_width: 3
    toc_depth: 2
always_allow_html: TRUE
urlcolor: blue
linkcolor: black
citecolor: blue
geometry: margin=1in
header-includes: 
  - \usepackage{titling}
  - \usepackage{fancyhdr}
  - \usepackage{graphicx}
  - \usepackage{float}
params:
  root_dir: './'
  metadata_dir: './'
  metadata_file: '.'
  PROJECT_NAME: '.'
  PI_NAME: '.'
  TASK_ID: '.'
  PROJECT_LEAD_NAME: '.'
  DEPARTMENT: '.'
  LEAD_ANALYSTS: '.'
  GROUP_LEAD: '.'
  CONTACT_EMAIL: '.'
  PIPELINE: '.'
  START_DATE: '.'
  COMPLETION_DATE: '.'
  data_dir: './'
  soup_fraction_value_default: '.'
---

```{r logo-file, echo=FALSE}
attach(params)
# Insert logo on the top of the html report 
logo_file <- file.path(root_dir, "figures", "img", "DNB-BINF-Core-logo.png")
htmltools::img(src = knitr::image_uri(logo_file), alt = "logo", style = "position:absolute; top:0; left:0; padding:0px; height:120px;")
detach(params)
```


\addtolength{\headheight}{2.0cm} 
\fancypagestyle{plain}{} 
\thispagestyle{fancy}
\fancyhead[L]{\includegraphics[height=120px]{`r logo_file`}}
\renewcommand{\headrulewidth}{0pt}

<style type="text/css">
:root {--DNB_BINF_Core_color: #00427B;}

h1.title {margin-top: 130px;
          margin-bottom: 25px;
          font-size: 36px;}

.nobullet li {list-style-type: none;}

.reporthead {font-size: 20px;}

body { /* Normal */
  font-size: 16px;
  font-style: Arial, Helvetica, sans-serif;}

h1 {color: var(--DNB_BINF_Core_color);
    font-size: 28px;
    margin-top: 50px;}

h2 {color: var(--DNB_BINF_Core_color);
    font-size: 20px;}

.list-group-item.active, .list-group-item.active:focus, .list-group-item.active:hover {
  background-color: var(--DNB_BINF_Core_color);}
</style>

<a href="https://wiki.stjude.org/display/CAB">

</a>

\pagebreak

<div class="reporthead"><br/>
**PI: `r params$PI_NAME`**  
**Project: `r params$PROJECT_NAME`**  
Task: `r params$TASK_ID`  
Project Lead(s): `r params$PROJECT_LEAD_NAME`  
Department: `r params$DEPARTMENT`  

<br />  

DNB Bioinformatics Core Analysis Team: 
<br />  

>**Lead Analyst(s): `r params$LEAD_ANALYSTS`**  
>Group Lead: `r params$GROUP_LEAD`  
<br />
>**Contact E-mail:** `r params$CONTACT_EMAIL`  
>**DNB Bioinformatics Core Pipeline:** `r params$PIPELINE`  

Date started: `r params$START_DATE`  
Date completed:  `r params$COMPLETION_DATE`  
Report generated: `r format(Sys.time(), '%H:%M:%S %Z %m/%d/%Y')` \

Reviewed by: _____________________   Date: ____________ \
</div>
\pagebreak

# Information about this notebook
[SoupX method](https://cran.r-project.org/web/packages/SoupX/vignettes/pbmcTutorial.html) profiles “the soup”, i.e., collection of cell-free mRNAs floating in the input solution. The soup looks different for each input solution and strongly resembles the expression pattern obtained by summing all the individual cells ([Young and Behjati, 2020](https://doi.org/10.1093/gigascience/giaa151)).

SoupX calculates `Cell-specific contamination fraction` (Estimate (or manually set) the contamination fraction, the fraction of UMIs originating from the background, in each cell) and infers a `corrected expression matrix` (Correct the expression of each cell using the ambient mRNA expression profile and estimated contamination).

The method to do this consists of three parts:

1. Calculate the profile of the soup.
2. Estimate the cell specific contamination fraction.
3. Infer a corrected expression matrix.


# Set up

We will set up libraries necessary for this notebook.

```{r load-library, echo=TRUE}
suppressPackageStartupMessages({
  library(future)
  library(knitr)
  library(SoupX)
  library(Seurat)
  library(stringr)
  library(tidyverse)
  library(tinytex)
  #library(SingleCellExperiment)
  library(Matrix)
  library(glue)
  
  # Evaluate Seurat R expressions asynchronously when possible using future package
  plan(multisession, workers = parallelly::availableCores())})
```

# Directories and paths to file Inputs/Outputs

We will set up directories and paths necessary for this notebook.


```{r set-dir-and-file-names, echo=TRUE}
attach(params)

analysis_dir <- file.path(root_dir, "analyses", "upstream-analysis") 
module_results_dir <- file.path(analysis_dir, paste0("results"))

# Input files
project_metadata_file <- file.path(metadata_dir, metadata_file) # metadata input file

# Create SoupX_results_dir
SoupX_results_dir <- file.path(module_results_dir, paste0("01_SoupX"))
if (!dir.exists(SoupX_results_dir)) {
  dir.create(SoupX_results_dir)}

# Create to plots directory
plots_dir <- file.path(analysis_dir, "plots") 

# Create SoupX_plots_dir
SoupX_plots_dir <- file.path(plots_dir, paste0("01_SoupX"))
if (!dir.exists(SoupX_plots_dir)) {
  dir.create(SoupX_plots_dir)}

source(paste0(root_dir, "/figures/scripts/theme_plot.R"))
```

```{r echo=FALSE,warning=FALSE}
opts_chunk$set(fig.align='center',
               external=TRUE,
               echo=FALSE,
               warning=FALSE,
               fig.pos='H',
               fig.path = paste0(SoupX_plots_dir, "/",  "SoupX-contamination-fraction-plot-", sep = ""))
a4width <- 8.3
a4height <- 11.7
```

# Read metadata file and define `sample_name`
```{r read-metadata-define-sample-name, echo=TRUE}
# Read metadata
project.metadata <- read.csv(project_metadata_file, sep = '\t', header = TRUE)
sample_name <- unique(as.character(project.metadata$ID))
sample_name <- sort(sample_name, decreasing = FALSE)
print(sample_name)
```

# Read in 10x data and run SoupX

We will estimate the contamination fraction `rho`. The posterior distribution is calculated using a Poisson likelihood with a gamma distribution prior, parametrised by its mean priorRho and standard deviation priorRhoStdDev. The dotted line in the above plot shows the prior distribution. The default parameters have been calibrated to be fairly non-specific with a slight preference towards values of rho in the 0% to 10% range which is most commonly seen for fresh (i.e. not nuclear) single cell experiments. The default values place only a very weak constraint, as can be seen by setting a uniform prior. 

If contamination fraction estimate is > 20%, it is recommended to set `rho` manually. In this case, we will set 5% background to be removed by default.

```{r run-SoupX, echo=TRUE}
# Create contamination summary table
#contamination_summary <- matrix(ncol = 4, nrow = 1) %>% as.data.frame()
#colnames(contamination_summary) <- c("sample_name", "rho_estimated_by_SoupX", "rho_adjusted", "rho_removed")
contamination_summary <- data.frame(Sample = character(),
                                    Estimated_Soup_Fraction = numeric(),
                                    Set_to_Default = character(),
                                    Default_Soup_Fraction = numeric(),
                                    stringsAsFactors = FALSE)

input_dir <- yaml$data_dir
cellranger_parameters <- yaml$cellranger_parameters

# Process each sample
multi_config_dirs <- Sys.glob(file.path(input_dir, cellranger_parameters, "multi_config_*"))

for (config_dir in multi_config_dirs) {
  per_sample_outs_dir <- file.path(config_dir, glue("multi_run_{cellranger_parameters}"), "outs", "per_sample_outs")
  samples_found <- list.dirs(per_sample_outs_dir, full.names = FALSE, recursive = FALSE)
  
  for (sample in samples_found) {
    cat("Beginning process for:", sample, "\n")
    sample_input_dir <- file.path(per_sample_outs_dir, sample)
    
    # Create directory 
    samples_SoupX_results_dir <- file.path(SoupX_results_dir, sample) 
    if (!dir.exists(samples_SoupX_results_dir)) dir.create(samples_SoupX_results_dir)
    print(samples_SoupX_results_dir)
    
    cat("Reading in data\n")
    filtered_matrix_file <- file.path(per_sample_outs_dir, sample, "count", "sample_filtered_feature_bc_matrix.h5") 
    raw_matrix_file <- file.path(per_sample_outs_dir, sample, "count", "sample_raw_feature_bc_matrix.h5") 
    clusters_file <- file.path(per_sample_outs_dir, sample, "count", "analysis", "clustering", "gene_expression_graphclust", "clusters.csv") 
    
    filtered_matrix <- Seurat::Read10X_h5(filtered_matrix_file, use.names = TRUE)
    raw_matrix <- Seurat::Read10X_h5(raw_matrix_file, use.names = TRUE)
    clusters <- read.csv(clusters_file)
    
    mDat <- data.frame(clusters = clusters$Cluster, row.names = clusters$Barcode)
    
    features_file <- file.path(per_sample_outs_dir, sample, "count", "sample_raw_feature_bc_matrix", "features.tsv.gz") 
    features <- read.delim(features_file, header = FALSE, stringsAsFactors = FALSE)
    colnames(features) <- c("gene_id", "gene_name", "feature_type")
    gene_expression_genes <- features$gene_name[features$feature_type == "Gene Expression"]
    
    common_genes <- intersect(rownames(raw_matrix), rownames(filtered_matrix))
    common_genes <- intersect(common_genes, gene_expression_genes)
    raw_matrix <- raw_matrix[common_genes, ]
    filtered_matrix <- filtered_matrix[common_genes, ]
    stopifnot(identical(rownames(raw_matrix), rownames(filtered_matrix)))
    
    sc_raw <- SoupChannel(raw_matrix, filtered_matrix, mDat)
    sc <- autoEstCont(sc_raw, forceAccept = TRUE)
    #sc <- autoEstCont(sc_raw, forceAccept = TRUE, tfidfMin = 0.1, soupQuantile = 0.9)
    print(sc)
    soup_fraction_value <- sc$fit$rhoEst
    
    if (soup_fraction_value > 0.2) {
      print(soup_fraction_value)
      cat("Estimated soup fraction >20%, set to 5% by default.\n")
      sc = setContaminationFraction(sc_raw, soup_fraction_value_default)
      contamination_summary <- rbind(contamination_summary,
        data.frame(
          Sample = sample,
          Estimated_Soup_Fraction = soup_fraction_value,
          Set_to_Default = "yes",
          Default_Soup_Fraction = soup_fraction_value_default
        )
      )
      print(sc)
    } else {
      cat("Estimated soup fraction is", soup_fraction_value, "\n")
      contamination_summary <- rbind(contamination_summary,
        data.frame(
          Sample = sample,
          Estimated_Soup_Fraction = soup_fraction_value,
          Set_to_Default = "no",
          Default_Soup_Fraction = NA
        )
      )
    }
    
    out <- adjustCounts(sc, roundToInt = TRUE)
    colnames(out) <- paste0(sample, ":", colnames(out))
    colnames(out) <- gsub('-1$', '', colnames(out))
    
    saveRDS(out, file.path(samples_SoupX_results_dir, "mtx_corrected_SoupX.rds"))
    cat("Processing complete for sample", sample, "\n")
  }
}

```

# Contamination fraction `rho` per sample
We will summarize the estimated and removed background due to contamination per sample.

```{r, fig.align = "left", results = "asis", message = FALSE, warning = FALSE, echo = FALSE}
tables1 <- contamination_summary
cat("  \n<div align=\"center\" style=\"font-size:80%\">  \n")
caption_value = "Summary of contamination fraction per sample"
print(knitr::kable(tables1, align = "lcccc", caption = caption_value))
cat("  \n</div>  \n")
cat("  \n\\pagebreak  \n")
```

# Save output files

```{r save-output, echo = TRUE}
#saveRDS(out_merge, file.path(SoupX_results_dir, "SoupX_corrected_mtx_merged.rds")) # Save combined matrices

contamination_summary %>%
  write_tsv(file.path(SoupX_results_dir,"SoupX_contamination_summary.tsv")) # Save `contamination_summary`
```

```{r echo=FALSE}
detach(params)
```

\pagebreak

# Session Info

```{r}
sessionInfo()
```

